# Introduction to `ggplot2`

## What is `ggplot2`

* The `ggplot2` package [@ggplot2] was developed to build a graphic from **few graphical components** (e.g., data, coordinate systems, geometric objects, aesthetics, facets, themes) based on the **grammar of graphics**. 

> "Wilkinson (2005) created the **grammar of graphics** to describe the deep features that underlie all statistical graphics. The grammar of graphics is an answer to a question: **what is a statistical graphic?** The **layered grammar of graphics** (Wickham, 2009) builds on Wilkinson’s grammar, focussing on the primacy of **layers** and adapting it for embedding within R. In brief, the grammar tells us that a statistical graphic is a **mapping** from **data** to **aesthetic attributes (colour, shape, size) of geometric objects (points, lines, bars)**. The plot may also contain **statistical transformations** of the data and is drawn on a specific **coordinate system**. **Faceting** can be used to generate the same plot for different subsets of the dataset. It is the combination of these independent components that make up a graphic."
>
> --- @ggplot2

* The components of `ggplot2`
    * Data
    * Geometric objects (**geom** for short)
    * Aesthetic mappings
    * Statistical transformations (**stats** for short)
    * Scales
    * A coordinate system (**coord** for short)
    * A faceting 

* Install the `tidyverse` package (or any package)
    * using `install.packages()` function
    * using the `Packages` pane 
    
```{r, eval = FALSE}
install.packages("tidyverse")
```
    
* Load the `tidyverse` package onto memory

```{r}
# We need to load a package whenever we use it
suppressMessages(library(tidyverse)) 
```


## Data

* "The **data** are what you want to visualise and a set of **aesthetic mappings** describe how **variables in the data** are mapped to **aesthetic attributes** that you can perceive." [@ggplot2]

```{r}
# diamonds is a built-in data in ggplot2
# ?diamonds display the help document for data 
# tibble is a datastructure in tidyverse 
diamonds
```


## Geometric objects (**geoms**)

* "Geometric objects, geoms for short, represent **what you actually see** on the plot: **points**, **lines**, **polygons**, etc." [@ggplot2]

```{r}
# ggplot() initializes a ggplot object.
# it can be used to specify 1) a dataset, and 2) aesthetic mapping
ggplot(data = diamonds, aes(x = carat, y = price))
```


```{r}
# geom_points() adds a new layer to a plot by drawing points to produce a scatter plot 
ggplot(data = diamonds, aes(x = carat, y = price)) + geom_point()
```

```{r}
# geom_smooth() adds an additional layer to the plot by drawing a smoothed line to capture the trend in the scatterplot
ggplot(data = diamonds, aes(x = carat, y = price)) + geom_point() + geom_smooth()
```



## Exercise 

* typing `mtcars` in your console will display the content of the `mtcars` dataset. How can we display the help document for the `mtcars` data?

* type `head(mtcars)`. What did `head()` do? Check the help document of `head()` (`mtcar` is a dataframe in base R, whereas the `diamonds` is a tibble in `tidyverse`). 

* Using the `mtcars` data, plot the scatter plot between `mpg` (miles per gallon: y axis) and `disp` (displacement: x axis) with a smoothed line. 

## Aesthetic mappings

* "A set of aesthetic mappings describe how **variables in the data** are mapped to **aesthetic properties** of the layer" [@ggplot2] 

* "To describe the way that **variables in the data** are mapped to **things that we can perceive on the plot (the "aesthetics")**, we use the `aes` function. The `aes` function takes a list of aesthetic-variable pairs like these: `aes(x = weight, y = height, colour = age)`. Here we are mapping x-position to `weight`, y-position to `height` and colour to `age`. The first two arguments can be left without names, in which case they correspond to the x and y variables." [@ggplot2]

```{r}
# color = color maps the variable 'color` in the dataset to the color aesthetics of points to encode further information in the graphic. 
ggplot(data = diamonds, aes(x = carat, y = price, color = color)) + geom_point()
```

```{r}
# shape = cut maps the variable 'cut` in the dataset to the shape aesthetics of points to encode further information in the graphic. 
# the graphic is not so informative because points are overplotted. Sometimes, facetting may handle overplotting 
ggplot(data = diamonds, aes(x = carat, y = price, shape = cut)) + geom_point()
```

```{r}
ggplot(data = diamonds, aes(x = carat, y = price)) + geom_point(color = "blue")
```



## Exercise 

* `mpg` is similar to `mtcars` but is a built-in tibble in `ggplot2`. 1) Plot `hwy` (mile per gallon: y axis) against `displ` (engine displancement: x axis), 2) Given the plot from 1), map the `class` variable to color, shape, alpha, and size aesthetics. 


* Explain what happens. 

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color=drv)) + geom_point() + geom_smooth(method="lm")
```

* This is what happens when mapping `hyw`, `displ`, and `cyl` to `x`, `y`, and `color` aesthetics. R creates a new dataset that contains all the data to be displayed on the plot. 

x    |y    |color
-----|-----|-----
1.8  |29   |4
1.8  |29   |4
2.0  |31   |4
2.0  |30   |4
2.8  |26   |6
2.8  |26   |6
3.1  |27   |6
1.8  |26   |4
1.8  |25   |4
2.0  |28   |4


## Scales 

* In the previous table, computers don't know how to display colors based on 4, 6, ... Computers need a a hexadecimal code for colors such as `FF6C91`. The mapping from the data to the final values that computers can use to display aesthetics is called a scale. 

* "The scales map values in the data space to values in an aesthetic space, whether it be colour, or size, or shape. Scales draw a legend or axes, which provide an inverse mapping to make it possible to read the original data values from the graph." [@ggplot2]

* `scale_x_continuous()` and `scale_y_continuous()` are the default scales for continuous x and y aesthetics: [https://ggplot2.tidyverse.org/reference/scale_continuous.html](https://ggplot2.tidyverse.org/reference/scale_continuous.html). 

```{r}
p1 <- ggplot(mpg, aes(displ, hwy)) + geom_point()
p1
```


```{r}
# change the axis labels
p1 + scale_x_continuous("Engine displacement (L)") +
  scale_y_continuous("Highway MPG")
```


```{r}
# also use the short-cut labs()
p1 + labs(x = "Engine displacement (L)", y = "Highway MPG")
```


```{r}
# modify the axis limits
p1 + scale_x_continuous(limits = c(2, 6))
```

```{r}
# use the short hand functions `xlim()` and `ylim()`
p1 + xlim(2, 6)
```

```{r}
#  choose where the ticks appear
p1 + scale_x_continuous(breaks = c(2, 4, 6))
```

```{r}
#  choose your own labels
p1 + scale_x_continuous(breaks = c(2, 4, 6), label = c("two", "four", "six"))
```


```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color=drv)) + geom_point() + geom_smooth(method="lm") + labs(title ="MPG vs Engine size", x = "Engine size", y = "MPG")
```


## Statistical transformations (**stats** for short)

```{r}
# historam shows the distribution of a single variable. 
# where does count come from? 
ggplot(data = diamonds, aes(x = carat)) + geom_histogram()
```

* "Statistical transformations, **stats** for short, summarise data in many useful ways. For example, binning and counting observations to create a histogram, or summarising a 2d relationship with a linear model. Stats are optional, but very useful." [@ggplot2]

* How `geom_histogram()` works? 
    * "A stat takes a dataset as input and returns a dataset as output, and so a stat can add new variables to the original dataset. It is possible to map aesthetics to these new variables. For example, `stat_bin`, the statistic used to make histograms, produces the following variables:
        * `count`, the number of observations in each bin
        * `density`, the density of observations in each bin (percentage of total / bar
width)
        * `x`, the centre of the bin" [@ggplot2]
    * "These generated variables can be used instead of the variables present in
the original dataset. For example, the default histogram geom assigns the height of the bars to the number of observations (`count`), but if you’d prefer a more traditional histogram, you can use the density (`density`). The following example shows a density histogram of carat from the diamonds dataset." [@ggplot2]

```{r}
# The names of generated variables must be surrounded with ..
 ggplot(diamonds, aes(carat)) + geom_histogram(aes(y = ..density..), binwidth = 0.1)
```

* Every **geom** has a default **stats**. 

* Position adjustments
    * Position adjustments determine how to arrange **geoms** that would otherwise occupy the same space. 

```{r}
# The discrete analogue of histogram is the bar plot
s <- ggplot(mpg, aes(fl, fill = drv))
```

```{r}
s + geom_bar()
```

```{r}
# Stack elements on top of one another
s + geom_bar(position = "stack")
``` 

```{r}
# Arrange elements side by side
s + geom_bar(position = "dodge")
```

```{r}
# Stack elements on top of one another,normalize height
s + geom_bar(position = "fill")
```

## A faceting 

* "A faceting specification describes how to break up the data into **subsets** and how to display those subsets as **small multiples**. This is also known as **conditioning** or latticing/trellising." [@ggplot2]

* "There are two types of faceting provided by ggplot2: `facet_grid` and `facet_wrap`. Facet grid produces a 2d grid of panels defined by variables which form the rows and columns, while facet wrap produces a 1d ribbon of panels that is wrapped into 2d" [@ggplot2]

```{r}
# facet into rows
ggplot(data = diamonds, aes(x = carat)) + geom_histogram() + facet_grid(color ~ .)
```

```{r}
# facet into columns
ggplot(data = diamonds, aes(x = carat)) + geom_histogram() + facet_grid(. ~ color)
```


## Exercise

* Using `mpg` data, plot `hwy` (y) vs `cty` (x). 

* facet into rows using `cyl`. 

* facet into columns using `cyl`. 

* facet into rows using `cyl` and columns using `year`


## Grouping

* "In many situations, you want to **separate your data into groups**, but render them in the same way. When looking at the data in aggregate you want to be able to distinguish individual subjects, but not identify them. This is common in **longitudinal studies** with many subjects, where the plots are often descriptively called spaghetti plots." [@ggplot2]

* `Oxboys` is a dataset in the `nlme` package. `Oxboys` includes the height of a selection of boys from Oxford, England versus a standardized age.

```{r}
library(nlme)
```

```{r}
# age = a numeric vector giving the standardized age 
head(Oxboys)
```

```{r}
ggplot(Oxboys, aes(age, height)) + geom_line()
```

```{r}
ggplot(Oxboys, aes(age, height, group = Subject)) + geom_line()
```

```{r}
# In many cases, this is not what we want
ggplot(Oxboys, aes(age, height, group = Subject)) + geom_line() + geom_smooth()
```

```{r}
# group = 1 override the default grouping 
ggplot(Oxboys, aes(age, height, group = Subject)) + geom_line() + geom_smooth(aes(group = 1))
```

```{r}
# facet is also useful for visualizing longitudinal data
ggplot(Oxboys, aes(age, height)) + geom_line() + facet_wrap(~Subject)
```

## Themes

"Themes are a powerful way to **customize** the non-data components of your plots: i.e. titles, labels, fonts, background, gridlines, and legends." More details are available at [https://ggplot2.tidyverse.org/reference/theme.html](https://ggplot2.tidyverse.org/reference/theme.html)

```{r}
ggplot(mpg, aes(x = hwy, y = cty)) + geom_point() 
```


```{r}
ggplot(mpg, aes(x = hwy, y = cty)) + geom_point() + theme(panel.background = element_rect(fill = "white", colour = "grey50"))
```


```{r}
ggplot(mpg, aes(x = hwy, y = cty)) + geom_point() + theme_classic()
```

## Save a ggplot

* `ggsave()` is a convenient function for saving a plot. It defaults to saving the **last plot** that you displayed. 

```{r}
ggsave("mtcars.pdf")
```

## More resources

* `ggplot2` Reference: [https://ggplot2.tidyverse.org/reference/index.html](https://ggplot2.tidyverse.org/reference/index.html)
* Many R galleries (e.g., [https://www.r-graph-gallery.com](https://www.r-graph-gallery.com))
* Google
