# Descriptive Statistics

## R functions for descriptive statistics


```{r}
# compute mean
# we use $ to access `price` variable in the `diamonds` dataset 
mean(diamonds$price)
```

```{r}
# compute median
median(diamonds$price)
```


```{r}
# compute variance
var(diamonds$price)
```

```{r}
# compute standard deviation
sd(diamonds$price)
```

```{r}
# summary of a data frame
summary(diamonds)
```

## qplot()

`qplot()`, short for **quick plot** is a function in the `ggplot2` package. qplot makes it easy to produce complex plots, often requiring several lines of code using other plotting systems, **in one line**.  

## Scatterplots

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) + geom_point()
```

```{r}
qplot(displ, hwy, data = mpg)
```

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, color = class)) + geom_point()
```

```{r}
qplot(displ, hwy, data = mpg, color = class)
```


```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, color = class, shape = drv)) + geom_point()
```

```{r}
qplot(displ, hwy, data = mpg, color = class, shape = drv)
```


```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) + geom_point() + geom_smooth(method = "lm")
```

```{r}
qplot(displ, hwy, data = mpg, geom = c("point", "smooth")) 
```

## Histogram

```{r}
ggplot(data = diamonds, aes(x = carat)) + geom_histogram()
```

```{r}
qplot(carat, data = diamonds, geom = "histogram")
```

```{r}
ggplot(data = diamonds, aes(x = carat)) + geom_histogram(binwidth = 0.05) + xlim(c(0,3))
```


```{r}
qplot(carat, data = diamonds, geom = "histogram", binwidth = 0.05, xlim = c(0,3))
```

```{r}
ggplot(data = mpg, aes(x = hwy, fill = drv)) + geom_histogram() 
```


```{r}
qplot(hwy, data = mpg, geom = "histogram", fill = drv)
```


## Density plots


```{r}
ggplot(data = diamonds, aes(x = carat, color = color)) + geom_density() 
```


```{r}
qplot(carat, data = diamonds, geom = "density", color = color)
```



## Barplots

```{r}
ggplot(data = diamonds, aes(x = clarity)) + geom_bar() 
```


```{r}
qplot(clarity, data = diamonds, geom = "bar")
```


```{r}
ggplot(data = diamonds, aes(x = clarity, fill = cut)) + geom_bar() 
```


```{r}
qplot(clarity, data = diamonds, geom = "bar", fill = cut)
```


## Boxplots

```{r}
ggplot(data = diamonds, aes(y = price)) + geom_boxplot() 
```

```{r}
qplot(y = price, data = diamonds, geom = "boxplot")
```

```{r}
ggplot(data = diamonds, aes(x = cut, y = price)) + geom_boxplot() 
```

## Faceting

```{r}
ggplot(data = diamonds, aes(x = carat, y = price)) + geom_point() + facet_grid(cut ~ color)
```

```{r}
qplot(carat, price, data = diamonds, facets = cut ~ color)
```


## The `corrplot` Package

* [https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html)
* "The corrplot package is a graphical display of a **correlation matrix**, confidence interval. It also contains some algorithms to do matrix reordering. In addition, corrplot is good at details, including choosing color, text labels, color labels, layout, etc."

```{r}
library(corrplot)
```

```{r}
corrplot.mixed(cor(mtcars))
```





