---
output:
  pdf_document: default
  html_document: default
---
# Base R Quiz


### Dataset1: `diamonds` is a dataset in `ggplot2` containing the prices and other attributes of almost 54,000 diamonds. In the dataset, `price` is the price of diamonds in US dollars, and `carat` is the weight of the diamonds (0.2-5.01). 

```{r}
library(ggplot2)
```

```{r}
diamonds
```

Let's create three vectors for `price`, `carat`, and `depth` variables in the `diamonds` dataset. 
```{r}
price <- diamonds$price
carat <- diamonds$carat
depth <- diamonds$depth
```

Now, we have a quite long vector. 
```{r}
# carat and depth should have the same length b/c they came from the same dataset. 
length(price)
```

Some quiz problems need the `round()` function, which will return the integer part of the provided number. **For quiz responses, please round your answers to the third decimal place.**

```{r}
# round to the zero digit (which is integer)
round(3.141592)
```
```{r}
# round to the first digit
round(3.141592, digits = 3)
```


### Quiz problem 1

Do we have any missing value in `price`? (Yes, or No)

```{r}
sum(is.na(price))
```

```{r}
sum(is.na(carat))
```

```{r}
sum(is.na(depth))
```

### Quiz problem 2

What is the sum of deviations of `price` from the mean?  

A deviation is computed by subtracting the mean of $X$ from a given score. $d_i=X_i- mean(X)$.

In other words, calculate $d_i = \sum_{i=1}^{n}[x_i-mean(x)]$, where $x$ is `price`. 

Report the integer part of the answer.  Use `round(x)` which returns $x$ without any decimals. 

```{r}
round(sum(price - mean(price)))
```


### Quiz problem 3

Variance is defined as average squared devaiton. 

$$Var(X)=\frac{\sum_{i=1}^n d_i^2}{n-1} =\frac{\sum_{i=1}^n (X-\bar{X})^2}{n-1}$$.

Numerator is the sum of squared deviations computed in the previous problem. The denominator equals ($n-1$). What is the variance of the `price` variable? 


```{r}
sum((price-mean(price))^2)/(length(price)-1)
```

```{r}
var(price)
```



### Quiz problem 4

You want to buy one of the diamonds in the `diamonds` dataset. However, you only(?) have $5,000. What is the largest size (in carat) of diamonds you can buy. 

```{r}
max(carat[price<=5000])
```


### Quiz problem 5

Conglaturation !!! You won Mega Millions (happy). Now, you don't have any budget limit. However, you have a strong belief that 7 is a lucky number. So you want to buy 7th, 77th, and 777th expensive diamonds. What is the total price of the three diamonds?   

```{r}
sum(sort(price, decreasing = TRUE)[c(7, 77, 777)])
```


### Dataset2: `bfi` is a dataset in the `psych` package containing 25 personality self report items taken from the International Personality Item Pool. In the dataset, 

  1. `gender` represents gender (Males = 1, Females = 2), 
  2. `education` represent educational attainment (1 = HS, 2 = finished HS, 3 = some college, 4 = college graduate 5 = graduate degree), 
  3. `age` represents age in years, and 
  4. `E3` is one of the five items that measures extraversion ("Know how to captivate people"). 

  

```{r}
## This is great- requires them to load and access data from a package.
# You need to install and load the `psych` package to use the `bfi` dataset.
library(psych)
```

Let's create three vectors for `gender`, `education`, and `age` variables in the `diamonds` dataset. 

```{r}
gender <- bfi$gender
education <- bfi$education
age <- bfi$age
E3 <- bfi$E3
```

### Quiz problem 6

What is the type of each of these vectors (chracter, integer, double, factor)?

```{r}
typeof(gender)
```

```{r}
typeof(education)
```

```{r}
typeof(age)
```

```{r}
typeof(E3)
```



### Quiz problem 7

What proportion of the sample is females?


```{r}
sum(gender == 2)
```

```{r}
sum(gender == 2)/length(gender)
```

```{r}
table(gender)
```

```{r}
prop.table(table(gender))
```


### Quiz problem 8

What is the sample mean difference in `E3` between males (gender = 1) younger than (<) 26 and females (gender = 2) younger than (<) 26? That is, mean of `E3` for males younger than 26 - mean of `E3` for females younger than 26. 

```{r}
mean(E3[gender==1 & age < 26], na.rm = TRUE) - mean(E3[gender==2 & age < 26], na.rm = TRUE)
```

### Quiz problem 9

In the `bfi` dataset, how many males with age 16 and highschool degree are there? 

```{r}
sum((gender == 1) & (education == 1) & (age == 16), na.rm = TRUE)
```


### Quiz problem 10


Using the `bfi` data set, what is the count of 42 year old females?

```{r}
sum(age == 42 & gender == 2)
```


