# Base R Practice

# Set 1

1. Calculate the sum of square of all the observations in the vector `a` from their mean. 

```{r}
# the original vector 
a <- c(1,2,3)
```

```{r}
# deviations from the mean
# a is a vector and mean(a) is a sinlge number
# notice R used recycling rule to match different lengths
a-mean(a)
```
```{r}
# square of deviations
(a-mean(a))^2
```

```{r}
# sum of square 
sum((a-mean(a))^2)
```


2. Without using R, calculate the variance of `a`.

```{r}
# variance = SS/(n-1), where SS = sum of square
# for a, n = 3
2/2
```

```{r}
# we can directly calculate variance using var()
var(a)
```

3. Without using R, what would be the fourth element of `v1+v2`?

```{r}
# 9 since R recycled the shorter vector v2 
# This is what happened: c(4,5,6,7) + c(10,2,10,2)
v1 <- c(4,5,6,7)
v2 <- c(10,2)
v1 + v2
```

4. Without using R, what would be the result?
```{r}
# sum() requires a numeric vector
# for that, TRUE is converted into 1, and FALSE is converted into 0
# essentially, this expression count "the number of TRUEs'
sum(c(TRUE, TRUE, FALSE, TRUE, FALSE))
```


5. In `mtcars` data, what is the mean (rounded to 2 decimal) of `mpg` for cars with 6 cylinders? 

```{r}
# $ is used to access a variable within a data.frame (or a list)
mpg <- mtcars$mpg
mpg
```
```{r}
cyl <- mtcars$cyl
cyl
```

```{r}
# this comparison operation produces a logical vector
cyl==6
```
```{r}
# logical indexing: only element corresponding to TRUE will be extracted
mpg[cyl==6]
```

```{r}
mean(mpg[cyl==6])
```

```{r}
# if you become familiar with R, then you can do this with oneline
mean(mtcars$mpg[mtcars$cyl==6])
```

6. Given the 1000000 random numbers from a standard normal distribution (a normal distribution with mean = 0 and sd = 1), what would be the proportion of random numbers greater than 0?


```{r}
# have you seen 1.96 before?
set.seed(777)
a <- rnorm(1000000) # equivalent to rnorm(1000, 0, 1)
length(a[a>0])/length(a)
```

```{r}
# alternative way to do the same thing
sum(a>0)/length(a)
```

7. Given the 1000000 random numbers from a standard normal distribution (a normal distribution with mean = 0 and sd = 1), what would be the proportion of random numbers greater than 1.96? 

```{r}
# have you seen 1.96 before?
set.seed(777)
a <- rnorm(1000000) # equivalent to rnorm(1000, 0, 1)
length(a[a>1.96])/length(a)
```

```{r}
# alternative way to do the same thing
sum(a>1.96)/length(a)
```


8. Without using R, what would be the first element of the following expression?

```{r}
# %in% operator 
# v1 %in% v2 returns a logical vector indicating 
# whether the elements of v1 are included in v2. 
c(1,2,3) %in% c(2,3,4,5,6)
```


9. Without using R, which element would be displayed first when displaying the level of the following factor?

```{r}
# "medium" will be displayed first according to alphabetical order
# notice this order does not represent intrinsic order in factor
# this order only applies to display or sort
# we use ordered() to create ordered factor
a <- factor(c("high", "high", "medium", "low"))
a
```

10. The following code will produce error (TRUE or FALSE)?

```{r eval=FALSE, include=FALSE}
# this code will produce error because a factor only allow some pre-specified values (i.e., levels)
a[1] <- "very high"
```




