---
output:
  html_document: default
  pdf_document: default
---

# Base R 2D Data Structure

```{r, include=FALSE}
tutorial::go_interactive()
```

## Matrices 

### What is a matrix? 
- A matrix is a two dimensional array of elements with rows and columns. 
- A matrix is **implemented as a vector** with special attributes
- Adding a **`dim()` attribute** to an atomic vector allows it to behave like a multi-dimensional array. 

```{r}
# create a vector a
a <- 1:6 
# attr() = get or set specific attributes of an object.
attr(a, "dim") <- c(3,2)
a
```

```{r}
a <- 1:6 
attr(a, "dim") <- c(3,2)
a
attributes(a)
```

```{r}
a <- 1:6 
attr(a, "dim") <- c(3,2)
a
dim(a)
```


### Create a matrix
- `matrix()` creates a matrix from the given set of values

```{r, echo=TRUE}
# by default, a matrix is filled by column
a <- matrix(1:6, ncol=3, nrow=2)
a
```

```{r}
# R will calculate nrow
a <- matrix(1:6, ncol=3)
a
```

```{r}
# a matrix can be filled by row using `byrow = TRUE`
a <- matrix(1:6, ncol=3, nrow=2, byrow = TRUE)
a
```

```{r, echo=TRUE}
# create an array of dimensions 
b <- array(1:12, c(2,3,2))
b
```

```{r, echo=TRUE}
b <- array(1:12, c(2,3,2))
b
dim(b)
```

-	`length()` generalises to `nrow()` and `ncol()` for matrices, and `dim()` for arrays.
-	`names()` generalises to `rownames()` and `colnames()` for matrices, and `dimnames()`, a list of character vectors, for arrays.

```{r, echo=TRUE}
results <- matrix(c(10, 30, 40, 50, 43, 56, 21, 30), 2, 4, byrow = TRUE)
colnames(results) <- c("1qrt", "2qrt", "3qrt", "4qrt")
rownames(results) <- c("store1", "store2")
results
```


## Data Frames

- A data frame is a **list** of **equal-length vectors**. 
- Since a list is a heterogeneous data structure, each element of a data frame can have different data types.  
- `data.frame()` creates a data frame. 

```{r, echo=TRUE}
# x = integer vector
# y = character vector
df <- data.frame(x=1:3, y=c("a", "b", "c"))
df
```

```{r, echo=TRUE}
# str() displays the internal structure of an R object
df <- data.frame(x=1:3, y=c("a", "b", "c"))
str(df) 
```

- `data.frame()` converts strings into factors by default. 
    - This default setting can cause serious problems in some cases. 
    - `stringAsFactors = FALSE` suppresses this default setting. 
    - Using `str()` to check data types is always a good practice. 


```{r, echo=TRUE}
df <- data.frame(x=1:3, y=c("a", "b", "c"), stringsAsFactors = FALSE)
str(df) 
```

```{r}
# a data frame is a list of equal-lengthed vectors
df <- data.frame(x=1:3, y=c("a", "b", "c"), stringsAsFactors = FALSE)
typeof(df)
```

## Operators

* Arithmetic Operators

|Operator	|Description |
|---------|------------|
|+        |addition    |
|-        |subtraction |
|*        |multiplication|
|/        |division|
|^ or **  |exponentiation|


* Logical Operators

|Operator	|Description |
|---------|------------|
| <       |less than    |
| <=       |less than or equal to |
|>        |greater than|
|>=       |greater than equal to|
|==  |exactly equal to|
|!=  |not equal to|
|!x  |Not x|
|x\|y  |x OR y|
|x&Y  |x AND y|


```{r}
1L == 2L
```

```{r}
"a" != "b"
```

### Quiz problem 1

(Without using R) What is the value of the following expression? 

`(1L == 2L) | ("a" != "b")`

A. TRUE

B. FALSE

### Quiz problem 2

(Without using R) What is the value of the following expression? 

`(1 == 2) & ("a" != "b")`

A. TRUE

B. FALSE


## Subsetting a Matrix and Dataframe

- **Positive integers** return elements at the specified positions.
- **Negative integers** omit elements at the specified positions
- **Logical vectors** select elements where the corresponding logical value is TRUE. 
- **Character vectors** return elements with matching names.
- **Nothing** returns the original vector. 

```{r, echo=TRUE}
# create a 3 x 3 matrix
a <- matrix(1:9, nrow = 3)
colnames(a) <- c("A", "B", "C")
a
```

```{r, echo=TRUE}
a <- matrix(1:9, nrow = 3)
colnames(a) <- c("A", "B", "C")

# If you omit any dimension, you obtain full columns or rows
# this is equivalent to `filter()` in tidyverse
a[2,]
```

```{r, echo=TRUE}
a <- matrix(1:9, nrow = 3)
colnames(a) <- c("A", "B", "C")

# If you omit any dimension, you obtain full columns or rows
# this is equivalent to `filter()` in tidyverse
a[c(1,2),]
```


```{r, echo=TRUE}
a <- matrix(1:9, nrow = 3)
colnames(a) <- c("A", "B", "C")

# this is equivalent to `select()` in tidyverse
a[,3]
```


```{r, echo=TRUE}
a <- matrix(1:9, nrow = 3)
colnames(a) <- c("A", "B", "C")

# this is equivalent to `select()` in tidyverse
a[,c("A", "B")]
```

```{r, echo=TRUE}
a <- matrix(1:9, nrow = 3)
colnames(a) <- c("A", "B", "C")

a[c(TRUE, FALSE, TRUE), c("B", "A")]
```

```{r}
a <- matrix(1:9, nrow = 3)
colnames(a) <- c("A", "B", "C")

a > 3
```


```{r, echo=TRUE}
a <- matrix(1:9, nrow = 3)
colnames(a) <- c("A", "B", "C")

a[a>3] <- NA
a
```

```{r}
# mtcars is a dataframe, and a dataframe is a list. 
# we use $ to access the element of a list. 
# therefore, we can use $ to access the column of a dataframe
str(mtcars)
```

```{r}
# accessing `mpg` variable in the `mtcars` dataset. 
mtcars$mpg
```

```{r}
# mpgb = a binary variable, 1 when mpg > 19, 0 otherwise
# This code creates a new binary variable `mpgb`. 
# This is equivalent to `mutate()` in tidyverse
mtcars$mpgb <- ifelse(mtcars$mpg > 19, 1, 0)
```



## `diamonds` dataset
- `diamonds` is a tibble, which is a tidyverse alternative to dataframe.
- We often subset the original dataset, and work on it. 

```{r}
library(ggplot2)
# subset only ideal cut diamonds
diamonds_ideal <- diamonds[diamonds$cut == "Ideal",]
# average price of ideal cut diamonds
mean(diamonds_ideal$price)
```

### Quiz problem 3

What is the average carat of "Premium" cut diamonds with "E" color? (2 decimal)

A. 0.52

B. 0.62

C. 0.72

D. 0.82

```{r}
library(ggplot2)
diamonds

# subset only premium cut diamonds with E color
diamonds_premium_e <- diamonds[diamonds$cut == "Premium" & diamonds$color == "E",]
# average price of ideal diamonds
mean(diamonds_premium_e$carat)
```


### Quiz problem 4

How many "Good" cut diamonds larger than 0.7 carat have "VS1" clarity?
What is the average carat of "Premium" diamonds with "E" color? (2 decimal)

A. 305

B. 587

C. 124

D. 987

```{r}
library(ggplot2)
diamonds

# subset only good cut diamonds larger than 0.7 carat
diamonds_good_0.7 <- diamonds[diamonds$cut == "Good" & diamonds$carat > 0.7,]
# average price of ideal diamonds
table(diamonds_good_0.7$clarity)
```


```{r}
library(psych)
bfi
```

## `bfi` dataset
- `bfi` is a dataframe in the `psych` package that contains 25 Personality items, sex, education, and age. 

### Quiz problem 5

How many missing values are in `education` in `bfi`? 

A. 123

B. 223

C. 323

D. 423

```{r}
library(psych)
sum(is.na(bfi$education))
```

### Quiz problem 6

Subset (i.e., filter()) `bfi` such that only `education` has no missing, and then calculate the frequency of male (coded as 1) using `table()`. 

A. 783

B. 694

C. 790

D. 838

```{r}
library(psych)
bfi
bfi_new <- bfi[!is.na(bfi$education), ]
table(bfi_new$gender)
```


### Quiz problem 7

Subset (i.e., filter()) `bfi` such that only males (coded as 1) are contained in the dataset, and then calculate the proportion of age 16 (round to 3 decimal). 

A. 0.013

B. 0.027

C. 0.033

D. 0.021

```{r}
library(psych)
bfi
bfi_male <- bfi[bfi$gender == 1, ]
table(bfi_male$age)/sum(table(bfi_male$age))
```






